---
title: 'Project 1: Exploratory Data Analysis'
author: "Caren Elhenawy"
date: '12-03-2020'
output:
  pdf_document:
    toc: no
  html_document:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F,
tidy=TRUE, tidy.opts=list(width.cutoff=60), R.options=list(max.print=100))
```

For this project, I chose to focus on mental health and the resources available in countries worldwide. The first dataset contains the suicide rates of 4 different years from each country. The suicide rates are for both genders combined and are age-standardized. The second dataset contains the amount of mental health specialitists present in the same countries during the year 2016. Each of the values in the second dataset, excluding the year, was per a population of 100000. I acquired both of these datasets from a website called Kaggle. As an individual who wants to be a psychiatrist, I was interested in seeing if there was a correlation between the amount of specialists (psychologists, nurses, and psychologists) to suicide rates. In particular, I wanted to see if the presence of specialists had decreased the suicide rates.

#Datasets:

```{r}
#Dataset1
suiciderates <- read.csv("suiciderates.csv")
#Dataset2
humanresources <- read.csv("Human Resources.csv")
```

#Tidying
```{r}
library("tidyverse")
library(tidyr)
library(dplyr)
```

#Tidying Dataset #1 
```{r}
suiciderates1 <- suiciderates%>%select(Country:X2000)%>%na.omit()
untidysuicide <- suiciderates1%>%pivot_wider(names_from = "Country", values_from = "X2016")
tidysuicide <- untidysuicide%>% pivot_longer(cols=c(4:186), names_to = "Country", values_to = "X2016")%>% na.omit()
tidysuicide
```

#Tidying Dataset #2 
```{r}
human1 <- humanresources%>%select(Country:Psychologists)
untidyhuman <- human1%>%pivot_wider(names_from = "Country", values_from = "Psychiatrists")
tidyhuman <- untidyhuman%>%pivot_longer(cols=c(5:111), names_to = "Country", values_to = "Psychiatrists")%>% na.omit()
tidyhuman
```

Since my datasets were originally tidy, I had to untidy them. I untidied both of my datastes using pivot_wider. For both datasets, I untidied them using a categorical (country) and a numeric (X2016 or Psychiatrists) variable . Following the untidying, I reordered my datasets using the function pivot_longer and omitted all NAs. 


#Joining 

```{r}
fulldata <- human1%>%full_join(suiciderates1)%>%na.omit()
fulldata
```

My two datasets were joined using the full_join function. I was interested in seeing all the numeric variables for each country. Some countries were dropped from the datatset since they did not have all variables. I was left to analyze data from 51 countries. 

#Wrangling
```{r}
fulldata%>%filter(between(Psychiatrists, 1, 17))
```

Using the filter function, I narrowed down the countries that had a psychiatrist rate between 1 and 17. I observed that only 25 countries out of the 51 had a psychiatrist rate greater than 1. The cut-off of 17 was chosen since all psychiatrists rates were below that. I was curious to see how many countries had a low amount of psychologists. I condluded that 26 out of the 51 countries had a psychologist rating of below 1. 

```{r}
fulldata%>%select(Country, Nurses)
```
Using the select fucntion, I paired the country and nurse column. I observed the amount of nurses in each country and noted that Turkey had the highest amount of nurses per a population of 100,000. Turkey had a nurse rate of 150.251 which is significanatly larger than the US rate of 4.283. 

```{r}
fulldata%>%arrange(desc(X2016))
```
Using the arrange function,I was able to see the descending suicide rates of 2016. Based on that table, the country Suriname had the highest suicide rate which was 23.2. 
```{r}
fulldata1 <- fulldata[!is.na(fulldata$Psychologists),]
fulldata1$lackofsupport <-  ifelse(fulldata1$Psychologists<median(fulldata1$Psychologists), "Low", "High")
lackofsupport <- fulldata1$lackofsupport
fulldata1
fulldata1%>%group_by(lackofsupport)
```

I created a new categorical variable called lackofsupport and used the group_by function to observe which countries had the lowest presence of psychologists.I chose the psychologist variable over other variables since psychologsists are the most common specialists to be visited for mental health concerns. Nurses and social workers are prominently visited for other reasons. Many countries that had a low presence of pscyhologists in 2016 had high suicide rates in that same year. For example, Uganda had a low psychologist rate in 2016 and the suicide rate in 2016 was 20.0. Azerbijan had a high psychologist rate of 1.165 and a suicide rate of 2.6 in the same exact year. There are exceptions to the aforementioned trend. One exception is Egypt. Egypt had a low psychologist rate of 0.256 and a low suicide rate of 4.4 in the exact same year. These exceptions indicate that there are other factors to be considered when assessing the cause of suicide and the effectiveness of mental health specialists. In regard to Egypt, since it is an extremely populated country, the psychologist rate might be low in comparison to its dense population. 

```{r}
fulldata2 <- fulldata1%>%mutate(averaage = Psychologists/Psychiatrists)
fulldata2
```

Using the mutate function, I created an additional column which divides the psychologist rate by the psychiatrist rate for each country. I wanted to see how many psyshologists there were in comparison to psychiatrists. Surprisingly, 23 out of the 51 countries had more psychiatrists than psycholgists. I expected there to be more psychologists since psychiatrists are typically visited only after the psychologist refers the patient to them. Psychiatrists are also only visited for their ability to perscribe medication. The high presence of psychiatrists suggests that there are more severe mental illnesses than expected. Severe mental illnesses require medication more often mild illnesses that can be handled by a psychologist.

```{r}
summary <- fulldata%>% summarize(mean_suicide2010 = mean(X2010, na.rm = T), n_country = n_distinct(Country))
summary
```

This summary shows the mean suicide rate of 2010 which was 10.14 and the amount of unique countries in the dataset is 51.

```{r}
fulldata1%>%summary()

fulldata1%>%group_by(Country, lackofsupport)%>%summarize(median(X2016))
fulldata1%>%group_by(Country, lackofsupport, Nurses)%>%summarize(mean(Psychiatrists))
fulldata1%>%summarize(mean(Psychologists), mean(Psychiatrists))
fulldata1%>%summarize(sd(Psychologists), sd(Psychiatrists))
fulldata1%>%summarize(max(Psychologists), max(Psychiatrists))
fulldata1%>%summarize(max(Social_workers), max(Nurses))
fulldata1%>%summarize(sd(X2015), sd(X2016))
fulldata1%>%summarize(sd(Social_workers), sd(Nurses))
fulldata1%>%summarize(n_distinct(Psychologists), n_distinct(Psychiatrists))
fulldata1%>%summarize(mean(X2000), mean(X2010))
```

This fulldata1 summary shows all the different numerical variables and their corresponding minimum, 1st quratile, median, mean, 3rd  quartile, and maximum values. Ten summary statistic were coded:
First Code - The first code was grouped by two categorical variables and the median suicide rate for 2016 was compared. I was able to see more clearly the effect that psychologists have on the suicide rates. There was no clear trend identifed. Some countries like Burkina Faso had an expected trend of a higher suicide rate when there was a low presence of psychologist. There were other countries that had similar suicide rates, but were high and low in the lackofsupport column. 
Second Code - The second code was grouped by lackofsupport and nurses and the mean psychiatrist rates were observed. I was interested to see if there were countries that had three high rates of specialities. I observed that countries like Bulgaria, Croatia, and Estonia had high rates of all specialties. This was interesting to note since all three countries are located in Eastern Europe. This prompted me to note that healthcare must be valued there since there is a surplus amount of individuals trained to help their community. 
Third Code - The third code compared the mean of pscyhologist and pscyhiatrist rate. The mean for psychologists was almost 3 times greater than psychiatrists, even though almost half of the countries observed had a greater psychiatrist rate than pscyhologist. The mean pscyhologist rate was 6.78 and the mean pscyhiatrist rate was 2.70. 
Fourth Code - The fourth code compared the standard deviations of pscyhologists and psychiatrists. The standard deviation for pscyhologists was 21.15, suggesting that there is a wide range of psychologist rates acorss the 51 countries. The standard deviation for pscyhiatrists is 3.88, suggesting that the range for the psychiatrist rate is much more narrow.
Fifth Code - The fifth code compared the maximum values of psychiatrist and psychologist rates. The maximum for pscyhologists was 142.018 and the maximum for pscyhiatrists was 16.188. 
Sixth Code - The sixth code tells me the maximum number of social workers and the maximum number of nurses. Nurses were almost double the amount of social workers.
Seventh Code - This compared the standard deviation of suicide rates from the year 2015 and 2016. The difference of standard deviations was very slim. It was 0.13 which makes sense since these two values were taken a year apart and the suicide rates would not have changed drastically. 
Eighth Code - This compared the standard devation of social workers and nurses. The standard deviation for rate of social workers was 13.506 and the standard deviation for the rate nurses was 24.46, suggesting that the range of the nurses rate is wider than the range for the social workers.
Ninth Code - The sixth code tells me that there are 49 unique psychologist rates and 50 unique psychiatrist rates in the dataset.
Tenth Code - This code compared the mean suicide rate in 2000 and the mean suicide rate in 2010. This was the most shocking statistic since the suicide rate decreased by only 1.1. I expected that over a decade and with the increased awareness of mental health, it would have been lowered more.

#Visualizations 
```{r}
library("ggplot2")
fulldata1%>%select_if(is.numeric)%>%cor%>%as.data.frame%>%rownames_to_column%>%
  pivot_longer(-1)%>%ggplot(aes(rowname,name,fill=value))+geom_tile()+ ggtitle("Correlation Heatmap of All Variables")+
  geom_text(aes(label=round(value,2)))+xlab("")+ylab("")+ scale_fill_gradient2(low="red", high="orange") + theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

This correlation heatmap depicts that many of my variables have weak correlations. One strong correlations is between social workers and psychologists. The correlation between these two specialists is 0.82. This means that as the abundance of social workers increase, the abundance of psychologists also increase. There are strong trends between the different years that suicide rates were measured. This is expected as there was no drastic change between the year 2000 to the year 2016 in any of the countries. I expected there is be some strong correlations between psychologists and suicide rates, but many are close to 0. They are negative correlations which signifies that sucide rates decrease as the abundance of psychologists increase.


```{r}
fulldata1%>%filter(Country=="Brazil"|Country=="Egypt"|Country=="Greece"|Country=="Japan"|Country=="Mexico")%>%
  ggplot(aes(x= Country, y= X2016, fill= Country))+
  geom_bar(stat="summary", fun.y="mean")+ 
  facet_wrap(~lackofsupport)+
  ggtitle("Graph 1: Suicide Rates Across 5 Different Continents") + 
  theme(axis.text.x = element_blank(), legend.position = "bottom") + scale_fill_brewer(palette = "OrRd") + scale_y_continuous(breaks=seq(0,20,1))+ ylab("Suicide Rates in the Year 2016")
```
In this graph, I chose 5 different countries that were each in a different continent. I intended to showcase all 7 continents, but my dataset only repersented 5. I wanted to observe the suicide rates in 2016 and their psychologist rates simultaneously. Despite the greater than median rate of psychologists in 2016, Japan had the highest rate of suicide in 2016 out of all the countries depicted. Furthermore, despite the greater or lower than median rate of psychologists, the suicide rates for the other 4 countries were relatively the same. The country that was the most different was Greece who had the lowest suicide rates and a greater than the median psychologist rate. This plot and prior calculations has made me question the effectiveness of specialists and the importance in accounting for other variables when examining the reason why an individual is sucidial and why they commit suicide. Not many of the statistics have clear trends. This is the most evident when observing the correlation heatmap above.


```{r}
fulldata1%>%ggplot(aes(x=Psychiatrists,y=Nurses))+geom_point(size=2, aes(color=X2016))+ scale_color_gradient(low="maroon", high="orange")+ ggtitle("Graph 2: Rate of Nurses and Psychiatrists in 51 Countries")+ scale_y_continuous(breaks = seq(0, 150,10))+ scale_x_continuous(breaks =  seq(0,20,5))+xlab("Psychiatrist Rate")
```
In this graph, the psychiatrist and nurses rates for all 51 countries in the dataset are mapped against each other. The suicide rates for the year 2016 are also mapped and the color changes based on the rate. At first glance, one notices the cluster of datapoints in the bottom left corner. Many of the countries have a psychiatrist rate below 2.5 and a nurse rate below 10. Comparing the suicide rates between those datapoints that have similar psychologist and psychiatrist rates, one notices that there is a great range of suicide rates. There is a datapoint that seems that the suicide rate is close to 20 and there is a datapoint that is right under that that has a suicide rate of below 5. This prompts me to again question the presence of health specialists and if they are as useful as one would hope. Many other factors should be taken in account to understand one's mental health and this graph has brought my attention to this. 

#Clustering
```{r}
library(cluster)
pam1 <- fulldata1%>%pam(k=3)
pam1
```


```{r}
pamcluster <- fulldata1%>% mutate(cluster=as.factor(pam1$clustering))
pamcluster %>% ggplot(aes(Social_workers, X2016, color = lackofsupport, shape = cluster))+ geom_point(size=2)+ ggtitle("Graph 3: Social Workers, Psyshcologist Rates, and Suicide Rates of 2016")+xlab("Social Workers Rate")+ylab("Suicide Rates in 2016")
```
In this cluster graph, the relationship between two numeric variables and one categorical variable based on a third numeric variable was observed. The greatest area of concentration is at the 0.0 to 5.0 rate for social workers. Within this area, there are a wide range of suicide rates. We can conclude only one plausiable conclusion which is the abundance of social workers per 100,000 is very low and is unable to have a signifncant affect on the population. From this  graph, you cannot conclude that the social workers are ineffective. There has to be more data involved. Also, based on the colors of the cluster, one can conclude that the majority of countries who have a low lackofsupport which means lower than the median rate of psychologists, have higher suicide rates. This is depicted by the teal datapoints that have a greater amount than the orange datapoints in the upper left of the graph. This graph contains one of the more evident trends throughout this project. 

```{r}
pam2 <- fulldata1%>%select(-lackofsupport,-Country)%>%pam(3)
pam2
```
```{r}
final <- fulldata1%>%mutate(cluster=as.factor(pam2$clustering))
final2 <- final%>%group_by(lackofsupport)%>%count(cluster)%>%arrange(desc(n))%>%
pivot_wider(names_from="cluster",values_from="n",values_fill = list('n'=0))
final2
```
```{r}
library(plotly)
final%>%plot_ly(x= ~Psychiatrists, y = ~X2016, z = ~Psychologists, color= ~cluster,
type = "scatter3d", mode = "markers") %>%
layout(autosize = F, width = 900, height = 400)
```
This clustering graph showcases three numeric variables on the graph. Each datapoint defines 1 of the 51 countries in the dataset. The clusters are defined by the three colors (orange, green, and purple). The most concentrated area is where suicide rates, psychiatrist rates, and psychologist rates are all below 5.0. There are some outliers outside of the concentrated area. One of the countries has a psychiatrist rate of 16.19, suicide rating of 14.4 in 2016, and psychologist rating of 6.46. 